---
title: "ERV-W family overview"
output:
  html_document:
    df_print: paged
---

I want to know what the composition of LTRs associated to ERV-W looks like in repeatmasker. What is the most common ERV-W LTR subfamily? 

I've considered ERV-W LTRs as reported by [Kojima, 2018](https://doi.org/10.1186/s13100-017-0107-y):

* LTR17
* LTR12
* LTR12B
* LTR12C
* LTR12D
* LTR12E
* LTR12F
* LTR30
* LTR35
* LTR35B
* MER41A
* MER41B
* MER41D
* MER41E
* MER41F
* MER41G

```{r setup, include=FALSE}
    knitr::opts_knit$set(root.dir = normalizePath("/Volumes/MyPassport/MN11_HERVW/invitro/")) 
```

```{r}
library(data.table)
library(tidyverse)
library(ggplot2)
library(ggpubr)

rmsk <- fread("./data/hg38_rmsk_TEtranscripts.bed", data.table = F)
colnames(rmsk) <- c("chr", "start", "end", "TE_id", "dot", "strand", "TE_info")

# To reduce the size...
rmsk <- rmsk %>% 
  filter(grepl("LTR", TE_info)) %>% 
  filter(grepl("ERV1", TE_info))

# From Kojima, 2018 PMID: 29308093
ervw_ltrs <- c("LTR17_dup",
                "LTR12_dup", 
                "LTR12B_dup", 
                "LTR12C_dup", 
                "LTR12D_dup", 
                "LTR12E_dup", 
                "LTR12F_dup",
                "LTR30_dup",
                "LTR35_dup", 
                "LTR35B_dup",
                "MER41A_dup", 
                "MER41B_dup", 
                "MER41D_dup", 
                "MER41E_dup", 
                "MER41F_dup", 
                "MER41G_dup")

ervw_ltrs_rmsk <- rmsk %>% 
  filter(grepl(paste(ervw_ltrs, collapse = "|"), TE_id))

ervw_ltrs_rmsk_summary <- ervw_ltrs_rmsk %>% 
  group_by(TE_info) %>% 
  summarise(count = n()) %>% 
  mutate(TE_info = fct_reorder(str_remove(pattern = ",ERV1,LTR", TE_info), desc(count)))

ervw_ltrs_rmsk_summary %>%
  ggplot(aes(x=TE_info, y=count)) + geom_bar(stat="identity") + 
  geom_text(aes(label=count), vjust=0) +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + labs(x="", y="Count") + 
  ggtitle("ERV-W associated LTRs", subtitle = " ")

# write.table(ervw_ltrs_rmsk, "./data/hg38_rmsk_TEtranscripts_ERVW_LTRs.bed", sep = "\t", quote = F, col.names = F, row.names = F)

ervw_ltrs_rmsk %>% 
  filter(grepl("LTR17", TE_id))
```

## How many are likely to be part of a provirus? 

In order to answer this question we intersected these LTRs to retrotector predictions (hg38) as:

`bedtools intersect -a ./data/hg38_rmsk_TEtranscripts_ERVW_LTRs.bed -b ./data/hg38_ERVs_prediction.bed -wo -f 0.5 > ./data/hg38_rmsk_TEtranscripts_ERVW_LTRs_intersect_retrotector.bed`

Namely, an LTR would be considered as part of a provirus if it overlaps >=50% of it's sequence with a prediction. 

```{r}
rmsk_ERVW_LTRs_intersect_retrotector <- fread("./data/hg38_rmsk_TEtranscripts_ERVW_LTRs_intersect_retrotector.bed", data.table = F, skip = 1)
colnames(rmsk_ERVW_LTRs_intersect_retrotector) <- c("chr", "start", "end", "TE_id", "dot", "strand", "TE_info",
                                                    "chr_ERV", "start_ERV", "end_ERV", "dot2", "strand_ERV", 
                                                    "dot3", "TE_id_ERV", "overlap")

# Length of the LTR
rmsk_ERVW_LTRs_intersect_retrotector$length_a <- rmsk_ERVW_LTRs_intersect_retrotector$end - rmsk_ERVW_LTRs_intersect_retrotector$start

# Percentage of LTR sequence overlapping
rmsk_ERVW_LTRs_intersect_retrotector$overlap_perc <- rmsk_ERVW_LTRs_intersect_retrotector$overlap / rmsk_ERVW_LTRs_intersect_retrotector$length_a

# Is it 100%?
rmsk_ERVW_LTRs_intersect_retrotector$overlap_fct <- ifelse(rmsk_ERVW_LTRs_intersect_retrotector$overlap_perc == 1, "100%", ">50%")

# How many overlap in the subfamily (total)? We need this to order the plot by descending order
totals <- rmsk_ERVW_LTRs_intersect_retrotector %>% 
  count(TE_info, name = "total")

# How many overlap >50% and how many 100%?
rmsk_ERVW_LTRs_intersect_retrotector_count <- rmsk_ERVW_LTRs_intersect_retrotector %>% 
  count(TE_info, overlap_fct, name = "count") %>% 
  left_join(totals, by = "TE_info") %>% 
  mutate(TE_info = fct_reorder(str_remove(TE_info, ",ERV1,LTR"), total, .desc = TRUE))

ggplot(rmsk_ERVW_LTRs_intersect_retrotector_count, aes(x = TE_info, y = count, fill = overlap_fct)) +
  geom_bar(stat = "identity") + 
  geom_text(
  aes(label = count),
  position = position_stack(vjust = 0.5)) +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + labs(x="", y="Count") + 
  ggtitle("ERV-W associated LTRs", subtitle = "Intersecting with a retrotector prediction")

# Keeping order from most frequent to least
te_order <- levels(rmsk_ERVW_LTRs_intersect_retrotector_count$TE_info)

# Add a label with the percentage based on the entire subfamily (from ervw_ltrs_rmsk_summary).
# In other words, what percentage of the LTR subfamily overlaps a provirus >50%, and what percentage of the LTR subfamily overlaps a provirus 100%?
rmsk_ERVW_LTRs_intersect_retrotector %>% 
  count(TE_info, overlap_fct, name = "count_retrotector") %>% 
  mutate(TE_info = str_remove(TE_info, ",ERV1,LTR"),
         TE_info_fct = factor(TE_info, levels = te_order)) %>% 
  left_join(ervw_ltrs_rmsk_summary, by = "TE_info") %>% 
  mutate(ratio = count_retrotector / count) %>% 
  mutate(overlap_fct = factor(overlap_fct, levels = c("100%", ">50%"))) %>% 
  ggplot(aes(x = TE_info_fct, y = count_retrotector, fill = overlap_fct)) +
  geom_bar(stat = "identity") + 
  geom_text(aes(label = sprintf("%.1f%%", ratio * 100)),
            position = position_stack(vjust = 0.5)) +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + labs(x="", y="Count") + 
  ggtitle("ERV-W associated LTRs", subtitle = "Intersecting with a retrotector prediction")
```






