---
title: "CRISPRa ERV-W differential expression analysis of genes"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
    knitr::opts_knit$set(root.dir = normalizePath("/Volumes/MyPassport/MN11_HERVW/invitro/")) 
```

Here we perform differential expression analysis (DEA) of gene expression for our CRISPRa experiments targeting ERV-W elements.

The experimental design is CRISPR activation on iPSC and day 30 iNGNs. There are two different guides (LV4009 and LV4070): LV4009 targets LTR17 (ERV-W), and LV4070 is designed to only target ERVW1 (syncytin). LV3470 serves as control targeting LacZ (gene not present in the human genome). There were several experimental batches, for which we account in DESeq2 model.

## Load libraries

Before we start, I load libraries and define some helper functions to visualize the effect as a mean plot (credits to Per Brattås! who wrote the original code for it)
```{r class.source = 'fold-hide'}
library(EnhancedVolcano)
library(data.table)
library(pheatmap)
library(tidyverse)
library(ggplot2)
library(DESeq2)
library(clusterProfiler)
library(org.Hs.eg.db)
library(dplyr)
library(ggpubr)
set.seed(10)
## getSignName ##
# Get significantly different gene names. 
# Taken from source code of the package deseqAbstraction which is no longer available on github.
# Credits to Per L. Brattås
# Parameters:
# x = results object from deseq
# p = padj threshold for significance
# l = log2FC threshold for significance
getSignName <- function(x,p,l=0) {
  up <- x[!is.na(x$padj) & x$padj < p & x$log2FoldChange > l,]
  down <- x[!is.na(x$padj) & x$padj < p & x$log2FoldChange < -l,]
  return(list(up=rownames(up),down=rownames(down)))
}
## getAverage ##
# Get average expression (normalized by median of ratios) of each of the conditions in a deseq object.
# Taken from source code of the package deseqAbstraction which is no longer available on github.
# Credits to Per L. Brattås
# Parameters:
# dds = deseq object
getAverage <- function(dds) {
  baseMeanPerLvl <- sapply( levels(dds$condition), function(lvl) rowMeans( counts(dds,normalized=TRUE)[,dds$condition == lvl] ) )
  baseSDPerLvl <- sapply( levels(dds$condition), function(lvl) apply( counts(dds,normalized=TRUE)[,dds$condition == lvl],1,sd ) )
  colnames(baseSDPerLvl) <- paste("st.dev:",colnames(baseSDPerLvl),sep="")
  return(list(Mean=baseMeanPerLvl,SD=baseSDPerLvl))
}

meanPlot_cus <- function(exp,test,c1 = "condition 1",c2 = "condition 2",p=.05,l=0,id=F, ttl="", 
                         repel=TRUE, col1="tomato", col2="darkturquoise", col3="grey", highlights=NA){
  sign <- getSignName(x = test,p = p,l = l)
  u <- sign$up
  d <- sign$down
  
  #color up and down sign..
  colVec <- ifelse(test = (rownames(exp) %in% u),
                   yes = col1,
                   no = ifelse(test = (rownames(exp) %in% d),
                               yes = col2, no =col3))
  colVec[is.na(colVec)] <- "steelblue" ## if NA make sure it's not counted as <p
  #size of points
  cexVec <- ifelse(test = (rownames(exp) %in% u),
                   yes = 0.35,
                   no = ifelse(test = (rownames(exp) %in% d),
                               yes = 0.35, no = 0.3))
  
  exp_log <- as.data.frame(log2(exp[,c(c1, c2)] + 0.5))
  exp_log$Name <- rownames(exp_log)
  
  exp_log$reg <- factor(ifelse(exp_log$Name %in% u, paste('upregulated in ', c1, ' (', length(u), ')', sep =''),
                               ifelse(exp_log$Name %in% d, paste('downregulated in ', c1,' (', length(d), ')', sep =''), paste('not significant', ' (', (nrow(test) - length(u) - length(d)), ')', sep=''))))
  
  library(ggrepel)
  if(repel == TRUE){
    plt <- ggplot(exp_log, aes(x=get(c2), y=get(c1), label=Name, color=reg)) + geom_point(aes(size=cexVec), alpha=0.7)+ scale_color_manual(values=c(col2, col3, col1))+ scale_size_continuous(range=c(1,2), guide="none")+ geom_text_repel(data = subset(exp_log, Name %in% u | Name %in% d),direction    = "y", nudge_y = 0.4, nudge_x = -0.5)
  }
  else{
    plt <- ggplot(exp_log, aes(x=get(c2), y=get(c1), color=reg)) + geom_point(aes(size=cexVec), alpha=0.7)+ scale_color_manual(values=c(col2, col3, col1))+ scale_size_continuous(range=c(1,2), guide="none")
  }
  plt <- plt + labs(x=paste("log2(mean ",c2,")",sep=""), 
                    y=paste("log2(mean ",c1,")",sep=""),
                    title=paste(ttl, paste(c1," vs. ",c2,sep=""), sep = ': '),
                    subtitle=paste("p-adj < ",p,", log2(fc) > ",l,sep=""))+ theme_bw() +
    theme( plot.title = element_text( size=14, face="bold"), text = element_text(size=15), panel.grid.minor = element_blank(), panel.grid.major = element_blank(), legend.title=element_blank()) 
  
  
  if(id==T) {
    
    identify(log2(exp[,1]),log2(exp[,2]),labels = rownames(exp))
    
  }
  
  if(!is.na(highlights)){
    plt <- plt + geom_point(data=exp_log[highlights,], aes(x=get(c2), y=get(c1)), colour="springgreen3", size=5, shape=1, stroke=2)
  }
  return(plt)
  
}
```


## Load data and metadata

Here we load the metadata of the samples (and split them by experiment and guide), and the TE count matrices (`te_counts` with all counts).

LV = lentivirus
experiment = Experimental batch
dpi = Days post induction

```{r}
samplesheet0 <- fread("./data/samplesheet_0dpi.tab", data.table = F, header = T)
samplesheet30 <- fread("./data/samplesheet_30dpi.tab", data.table = F, header = T)
samplesheet <- rbind(samplesheet0, samplesheet30)

samplesheet_list <- split(samplesheet, f = list(samplesheet$LV, samplesheet$experiment, samplesheet$dpi))

samplesheet_comparisons <- list("Ca_LV4009_E0010405061213_0dpi" = rbind(samplesheet_list$LV4009.E001.0, 
                                                                        samplesheet_list$LV4009.E004.0,
                                                                        samplesheet_list$LV4009.E005.0,
                                                                        samplesheet_list$LV4009.E006.0,
                                                                        samplesheet_list$LV4009.E012.0,
                                                                        samplesheet_list$LV4009.E013.0,
                                                                        samplesheet_list$LV3470.E001.0, 
                                                                        samplesheet_list$LV3470.E004.0,
                                                                        samplesheet_list$LV3470.E005.0,
                                                                        samplesheet_list$LV3470.E012.0,
                                                                        samplesheet_list$LV3470.E013.0,
                                                                        samplesheet_list$LV3470.E006.0),
                                "Ca_LV4070_E00713_0dpi" = rbind(samplesheet_list$LV4070.E007.0,
                                                                samplesheet_list$LV3470.E007.0,
                                                                samplesheet_list$LV4070.E013.0,
                                                                samplesheet_list$LV3470.E013.0),
                                "Ca_LV4009_E01213_30dpi" = rbind(samplesheet_list$LV4009.E012.30,
                                                                 samplesheet_list$LV4009.E013.30,
                                                                 samplesheet_list$LV3470.E012.30,
                                                                 samplesheet_list$LV3470.E013.30))
path <- "~/inbox/MN11_HERVW/gene_counts/unique/"
samples <- samplesheet$sample
files <- list.files(path)
files <- files[which(grepl("gene_count", files))]
# Check which samples are missing
missing_samples <- samples[!sapply(samples, function(s) {
  any(grepl(s, files))
})]

missing_samples

for(i in 1:length(samples)){
  sample <- samples[i]
  if(i == 1){
    gene_counts <- fread(paste(path, sample, "_gene_count_matrix_2.csv", sep=""), data.table = F)    
    colnames(gene_counts)[ncol(gene_counts)] <- sample
    rownames(gene_counts) <- gene_counts$Geneid
    row_order <- rownames(gene_counts)
  }else{
    tmp <- fread(paste(path, sample, "_gene_count_matrix_2.csv", sep=""), data.table = F)
    colnames(tmp)[ncol(tmp)] <- sample
    rownames(tmp) <- tmp$Geneid
    gene_counts <- cbind(gene_counts[row_order,], tmp[row_order,sample,drop=F])
  }
}
```

## Gene Differential Expression Analysis (DEA)

Four comparisons to be made:

* LTR17 CRISPRa vs Control in hiPSC
* Syncytin CRISPRa vs Control in hiPSC
* LTR17 CRISPRa vs Control in day 30 iNGNs
* Syncytin CRISPRa vs Control in day 30 iNGNs
```{r}
gene_dds_list <- list()
gene_res_list <- list()
gene_res_list_df <- list()
gene_exp_list <- list()
rownames(gene_counts) <- gene_counts$Geneid

for(i in names(samplesheet_comparisons)){
  rownames(samplesheet_comparisons[[i]]) <- samplesheet_comparisons[[i]]$sample
  print(i)
  conditions <- unique(samplesheet_comparisons[[i]]$condition)
  print(c("Conditions: ", conditions))

  effect <- conditions[which(conditions != "Control")]
  test <- paste("condition_", effect ,"_vs_Control", sep = "")
  print(c("Test: ", test))
  counts <- gene_counts[,rownames(samplesheet_comparisons[[i]])]
  gene_dds_list[[i]] <- DESeqDataSetFromMatrix(counts, samplesheet_comparisons[[i]], design =  ~ experiment+condition)
  gene_dds_list[[i]]$condition <- relevel(gene_dds_list[[i]]$condition, "Control")
  gene_dds_list[[i]] <- DESeq(gene_dds_list[[i]])
  gene_res_list[[i]] <- lfcShrink(gene_dds_list[[i]], test)
  gene_exp_list[[i]] <- getAverage(gene_dds_list[[i]])
  gene_res_list_df[[i]] <- as.data.frame(gene_res_list[[i]])
  gene_res_list_df[[i]]$gene_name <- rownames(gene_res_list_df[[i]])
}

# And one more comparison which was done in one experimental batch
i <- "Ca_LV4070_E01213_30dpi" 
samplesheet_Ca_LV4070_E01213_30dpi <- rbind(samplesheet_list$LV4070.E013.30,
                                            samplesheet_list$LV3470.E013.30)
rownames(samplesheet_Ca_LV4070_E01213_30dpi) <- samplesheet_Ca_LV4070_E01213_30dpi$sample
counts <- gene_counts[,rownames(samplesheet_Ca_LV4070_E01213_30dpi)]
gene_dds_list[[i]] <- DESeqDataSetFromMatrix(counts, samplesheet_Ca_LV4070_E01213_30dpi, design =  ~ condition)
gene_dds_list[[i]]$condition <- relevel(gene_dds_list[[i]]$condition, "Control")
gene_dds_list[[i]] <- DESeq(gene_dds_list[[i]])
gene_res_list[[i]] <- lfcShrink(gene_dds_list[[i]], test)
gene_exp_list[[i]] <- getAverage(gene_dds_list[[i]])
gene_res_list_df[[i]] <- as.data.frame(gene_res_list[[i]])
gene_res_list_df[[i]]$gene_name <- rownames(gene_res_list_df[[i]])
```

## PCA plots

```{r}
gene_vst_list <- list()
for (i in names(gene_dds_list)) gene_vst_list[[i]] <- varianceStabilizingTransformation(gene_dds_list[[i]])

pca_plots <- list()
for(i in names(gene_vst_list)) {
  pca_plots[[i]] <- plotPCA(gene_vst_list[[i]]) + ggtitle(i) + theme_bw()
}

pca_plots
```

## Gene markers 

Pluripotency, neuronal, ectoderm, endoderm, mesoderm, trophoectoderm markers

```{r}
gene_markers <- c("POU5F1", "NANOG", "SOX2", # ipsc
  "NEUROD4", "NEUROG2", "FOXG1", "NEUROD1", "SOX1",
  "HES5", "PAMR1", "PAX6",
  "CER1", "EOMES", "GATA6", 
  "APLNR", "HAND1", "HOXB7",
  "GATA3", "PTGES", "PDGFA",
  "MAP2", "DCX", "NCAM1") # Neuronal

# Normalize counts from hiPSCs
gene_counts_norm_hipsc <- gene_counts[, samplesheet_comparisons$Ca_LV4009_E0010405061213_0dpi$sample ] 
gene_counts_norm_hipsc[] <- mapply(gene_counts_norm_hipsc, gene_dds_list$Ca_LV4009_E0010405061213_0dpi$sizeFactor,  FUN="/")
rownames(gene_counts_norm_hipsc) <- rownames(gene_counts)
# Only LacZs
pheatmap(log2(gene_counts_norm_hipsc[gene_markers,samplesheet_comparisons$Ca_LV4009_E0010405061213_0dpi[which(samplesheet_comparisons$Ca_LV4009_E0010405061213_0dpi$condition == "Control"),"sample"]]+1), cluster_rows = F, show_colnames = F)
# Whole experiment 13 (example of CRISPRa hiPSCs too)
pheatmap(log2(gene_counts_norm_hipsc[gene_markers,samplesheet_comparisons$Ca_LV4009_E0010405061213_0dpi[which(samplesheet_comparisons$Ca_LV4009_E0010405061213_0dpi$condition == "Control" & samplesheet_comparisons$Ca_LV4009_E0010405061213_0dpi$experiment == "E013"),"sample"]]+1), cluster_rows = F, show_colnames = F)

# Normalize counts from neurons
gene_counts_norm_day30 <- gene_counts[, samplesheet_comparisons$Ca_LV4009_E01213_30dpi$sample ] 
gene_counts_norm_day30[] <- mapply(gene_counts_norm_day30, gene_dds_list$Ca_LV4009_E01213_30dpi$sizeFactor,  FUN="/")
rownames(gene_counts_norm_day30) <- rownames(gene_counts)

# Only LacZs
pheatmap(log2(gene_counts_norm_day30[gene_markers,samplesheet_comparisons$Ca_LV4009_E01213_30dpi[which(samplesheet_comparisons$Ca_LV4009_E01213_30dpi$condition == "Control"),"sample"]]+1), cluster_rows = F, show_colnames = F)
# Whole experiment 13 (example of CRISPRa day 30 iNGNs too)
pheatmap(log2(gene_counts_norm_day30[gene_markers,samplesheet_comparisons$Ca_LV4009_E01213_30dpi[which(samplesheet_comparisons$Ca_LV4009_E01213_30dpi$condition == "Control" & samplesheet_comparisons$Ca_LV4009_E01213_30dpi$experiment == "E013"),"sample"]]+1), cluster_rows = F, show_colnames = F)
```

## Mean plots, volcano plots and MA plots

To visualize gene DEA

```{r}
volcano_plots_list <- list()
mean_plots_list <- list()
for(i in names(gene_res_list)){
  tmp <- EnhancedVolcano(gene_res_list[[i]], 
                                             drawConnectors = T, 
                                             lab = NA, 
                                             x = 'log2FoldChange',
                                             y = 'padj',
                                             pCutoff = 0.05,
                                             FCcutoff = 1)
  tmp$data$Sig <- ifelse(tmp$data$log2FoldChange > 1 & tmp$data$Sig == "FC_P", "Upregulated", 
                                             ifelse(tmp$data$log2FoldChange < -1 & tmp$data$Sig == "FC_P", "Downregulated", "Not significant"))
  tmp$data$Sig <- paste(tmp$data$Sig, table(tmp$data$Sig)[tmp$data$Sig], sep = ": ")
  
  not_sig_lab <- unique(tmp$data$Sig)[startsWith(unique(tmp$data$Sig), "Not significant")]
  up_sig_lab <- unique(tmp$data$Sig)[startsWith(unique(tmp$data$Sig), "Upregulated")]
  down_sig_lab <- unique(tmp$data$Sig)[startsWith(unique(tmp$data$Sig), "Downregulated")]
  
  point_colours <- c("grey","darkturquoise", "tomato")
  names(point_colours) <- c(not_sig_lab, down_sig_lab, up_sig_lab)
  point_sizes <- c(1,2,2)
  names(point_sizes) <- c(not_sig_lab, down_sig_lab, up_sig_lab)  
  
  title <- str_replace_all(i, pattern = "_", replacement = " ")
  
  tmp <- tmp$data
  volcano_plots_list[[i]] <- ggplot(tmp, aes(x=log2FoldChange, y=-log10(padj), colour=Sig, size=Sig)) + geom_point(alpha=0.7) + 
    geom_vline(xintercept = 1, linetype="dotted") + geom_vline(xintercept = -1, linetype="dotted") + 
    geom_hline(yintercept = -log10(0.05), linetype="dotted") + labs(colour="") + 
    scale_color_manual(values = point_colours)  + 
    scale_size_manual(values = point_sizes, guide="none") +
    theme_bw() +
    theme(text = element_text(size=15), panel.grid.minor = element_blank(), panel.grid.major = element_blank()) +
    ggtitle(title)
  
  effect <- "CRISPRa"
  mean_plots_list[[i]] <- meanPlot_cus(exp = gene_exp_list[[i]]$Mean, 
                                       test = gene_res_list[[i]], 
                                       c1 = effect, c2="Control", p = 0.05, l=1, ttl = title, repel = F)
    
}

volcano_plots_list
mean_plots_list
```

## Gene Set Enrichment Analysis

We don't see anything major on the gene expression, yet, I wonder if there is any minor trend.

```{r}
gse_list <- list()
for(i in names(gene_res_list_df)){
  genelist <- gene_res_list_df[[i]][,c("log2FoldChange"), drop=F]
  genelist$gene_name <- rownames(genelist)
  genelist <- genelist[order(genelist$log2FoldChange, decreasing = T),]
  genelist_FC <- genelist$log2FoldChange
  names(genelist_FC) <- genelist$gene_name
  
  gse_list[[i]] <- gseGO(geneList=sort(genelist_FC, decreasing = T), 
               ont ="BP", 
               keyType = "SYMBOL", 
               minGSSize = 3, 
               maxGSSize = 800, 
               pvalueCutoff = 0.05,
               verbose = TRUE, 
               seed = TRUE,
               OrgDb = org.Hs.eg.db, 
               pAdjustMethod = "BH")
}

top20_gsea_dotplot <- function(gse, name){
  gse@result$sign <- ifelse(gse@result$enrichmentScore > 0, "Activated", "Supressed")
  gse@result$num_genes <- sapply(str_split(gse@result$core_enrichment, "/"), length)
  gse@result$gene_ratio <- gse@result$num_genes / gse@result$setSize
  
  gse@result <- top_n(gse@result, -20, gse@result$p.adjust) %>% arrange(gene_ratio)
  if(nrow(gse@result) > 20){
    gse@result <- top_n(gse@result, 20, abs(gse@result$enrichmentScore)) %>% arrange(gene_ratio)
  }
  gse@result$Description <- factor(gse@result$Description, levels = unique(gse@result$Description))
  
  ggplot(gse@result, aes(x=Description, y=gene_ratio, size = num_genes, fill = p.adjust)) + geom_point(shape=21, colour="lightgrey") + facet_wrap(.~sign) + coord_flip() + theme_pubr(legend = "right", border = T) + labs(x="", fill="Padj", y = "Gene ratio", size = "Num genes") + scale_fill_gradientn(colors = c("firebrick1", "gray94")) + scale_size_continuous(range = c(2,8)) + 
    theme(axis.text.y = element_text(size=10), strip.text.x = element_text(size = 10),axis.text.x = element_text(size=10),legend.text = element_text(size=10),legend.title = element_text(size=10)) + 
    ggtitle(paste(name, ":\nTop 20 most significant"))
}

top20_gsea <- list()
for( g in names(gse_list)){
  print(g)
  print(nrow(gse_list[[g]]@result))
  if(nrow(gse_list[[g]]@result) > 0){
    top20_gsea[[g]] <- top20_gsea_dotplot(gse = gse_list[[g]], name=g)  
  }
}

top20_gsea

# Get those results into excel files
gse_list_df <- list()
for(i in names(gse_list)){
  gse_list_df[[i]] <- as.data.frame(gse_list[[i]]@result)
  if(nrow(gse_list_df[[i]]) > 1){
    gse_list_df[[i]]$sign <- ifelse(gse_list_df[[i]]$enrichmentScore > 0, "Activated", "Supressed")
    gse_list_df[[i]]$num_genes <- sapply(str_split(gse_list_df[[i]]$core_enrichment, "/"), length)
    gse_list_df[[i]]$gene_ratio <- gse_list_df[[i]]$num_genes / gse_list_df[[i]]$setSize
    
    rownames(gse_list_df[[i]]) <- gse_list_df[[i]]$Description
    gse_list_df[[i]]$Description <- factor(gse_list_df[[i]]$Description, levels= gse_list_df[[i]][order(gse_list_df[[i]]$gene_ratio, decreasing = F),"Description"])
  }
}

openxlsx::write.xlsx(gse_list_df, "./data/genes_GSEA.xlsx")
openxlsx::write.xlsx(gene_res_list_df, "./data/genes_DEA.xlsx")
save.image("./data/gene_DEA.Rdata")
```

## Core enrichment gene examples

Would be good to show the size of the effect on the different experiments. Barplot of the meand and standard error of some of the genes listed as core enrichment of these terms.

```{r}
# Set gene names as rownames
gene_res_list_df$Ca_LV4009_E01213_30dpi$gene <- rownames(gene_res_list_df$Ca_LV4009_E01213_30dpi)

barplot_gene_e01213_dpi30 <- function(gene){
  return(gene_counts_norm_day30[gene,] %>% 
           rownames_to_column("gene") %>% 
           pivot_longer(-gene) %>% 
           rename(sample = name) %>% 
           left_join(samplesheet_comparisons$Ca_LV4009_E01213_30dpi) %>% 
           left_join(gene_res_list_df$Ca_LV4009_E01213_30dpi) %>% 
           ggplot(aes(x=condition, fill = condition, y=value)) + 
           stat_summary(fun = median, geom="bar") +
           stat_summary( fun.data = mean_se, geom = "errorbar", width=0.2) +
           geom_jitter(width = 0.2, alpha=0.4) +
           labs(y="Normalized Expression", x="") +
           ggtitle(gene, subtitle = paste0("log2FC=", format(gene_res_list_df$Ca_LV4009_E01213_30dpi[gene, "log2FoldChange"], digits=2),
                                           ", p-value=", format(gene_res_list_df$Ca_LV4009_E01213_30dpi[gene, "pvalue"], digits=2),
                                           "\np-adj=", format(gene_res_list_df$Ca_LV4009_E01213_30dpi[gene, "padj"], digits=2))) +
           theme_bw() + theme(legend.position = "None"))
}

barplot_gene_e01213_dpi30("GIT1")
barplot_gene_e01213_dpi30("PINK1")
barplot_gene_e01213_dpi30("VGF")
barplot_gene_e01213_dpi30("SLC25A22")
barplot_gene_e01213_dpi30("SIRT3")
barplot_gene_e01213_dpi30("NDUFA11")
barplot_gene_e01213_dpi30("NDUFA2")
barplot_gene_e01213_dpi30("NDUFB11")
barplot_gene_e01213_dpi30("NDUFB2")
barplot_gene_e01213_dpi30("NDUFB4")
barplot_gene_e01213_dpi30("NDUFB6")
barplot_gene_e01213_dpi30("NDUFB7")
barplot_gene_e01213_dpi30("NDUFS6")
barplot_gene_e01213_dpi30("NDUFS7")
barplot_gene_e01213_dpi30("NDUFS8")
barplot_gene_e01213_dpi30("NDUFV1")
barplot_gene_e01213_dpi30("AKT1")
barplot_gene_e01213_dpi30("PPP1R1A")
barplot_gene_e01213_dpi30("PPP1CA")
```

## Is it syncytin? 

We can compare ERV-W CRISPRa with Syncytin CRISPRa on day 30 iNGNs

```{r}
gene_counts_norm_syn_day30 <- gene_counts[, samplesheet_Ca_LV4070_E01213_30dpi$sample ] 
gene_counts_norm_syn_day30[] <- mapply(gene_counts_norm_syn_day30, gene_dds_list$Ca_LV4070_E01213_30dpi$sizeFactor,  FUN="/")
rownames(gene_counts_norm_syn_day30) <- rownames(gene_counts)

gene_res_list_df$Ca_LV4070_E01213_30dpi$gene <- rownames(gene_res_list_df$Ca_LV4070_E01213_30dpi)

barplot_gene_e013_dpi30 <- function(gene){
  return(gene_counts_norm_syn_day30[gene,] %>% 
           rownames_to_column("gene") %>% 
           pivot_longer(-gene) %>% 
           rename(sample = name) %>% 
           left_join(samplesheet_Ca_LV4070_E01213_30dpi) %>% 
           left_join(gene_res_list_df$Ca_LV4070_E01213_30dpi) %>% 
           ggplot(aes(x=condition, fill = condition, y=value)) + 
           stat_summary(fun = median, geom="bar") +
           stat_summary( fun.data = mean_se, geom = "errorbar", width=0.2) +
           geom_jitter(width = 0.2, alpha=0.4) +
           labs(y="Normalized Expression", x="") +
           ggtitle(gene, subtitle = paste0("log2FC=", format(gene_res_list_df$Ca_LV4070_E01213_30dpi[gene, "log2FoldChange"], digits=2),
                                           ", p-value=", format(gene_res_list_df$Ca_LV4070_E01213_30dpi[gene, "pvalue"], digits=2),
                                           "\np-adj=", format(gene_res_list_df$Ca_LV4070_E01213_30dpi[gene, "padj"], digits=2))) +
           theme_bw() + theme(legend.position = "None"))
}

barplot_gene_e013_dpi30("GIT1")
barplot_gene_e013_dpi30("PINK1")
barplot_gene_e013_dpi30("VGF")
barplot_gene_e013_dpi30("SLC25A22")
barplot_gene_e013_dpi30("SIRT3")
barplot_gene_e013_dpi30("NDUFA11")
barplot_gene_e013_dpi30("NDUFA2")
barplot_gene_e013_dpi30("NDUFB11")
barplot_gene_e013_dpi30("NDUFB2")
barplot_gene_e013_dpi30("NDUFB4")
barplot_gene_e013_dpi30("NDUFB6")
barplot_gene_e013_dpi30("NDUFB7")
barplot_gene_e013_dpi30("NDUFS6")
barplot_gene_e013_dpi30("NDUFS7")
barplot_gene_e013_dpi30("NDUFS8")
barplot_gene_e013_dpi30("NDUFV1")
barplot_gene_e013_dpi30("AKT1")
barplot_gene_e013_dpi30("PPP1R1A")
barplot_gene_e013_dpi30("PPP1CA")
```

## Correlation of energy related genes between ERV-W CRISPRa and syncytin CRISPRa

Are ANY of these core enrichment genes up in the syncytin CRISPRa?

```{r}
gene_res_list_df$Ca_LV4009_E01213_30dpi[unique(unlist(str_split(gse_list_df$Ca_LV4009_E01213_30dpi$core_enrichment, "/"))),] %>% 
  rename("LV4009_log2FC" = log2FoldChange) %>% 
  dplyr::select(LV4009_log2FC, gene_name) %>% 
  left_join(gene_res_list_df$Ca_LV4070_E01213_30dpi[unique(unlist(str_split(gse_list_df$Ca_LV4009_E01213_30dpi$core_enrichment, "/"))), ]) %>% 
  rename("LV4070_log2FC" = log2FoldChange) %>% 
  dplyr::select(LV4009_log2FC, LV4070_log2FC, gene_name) %>% 
  ggplot(aes(x=LV4009_log2FC, y=LV4070_log2FC, label=gene_name)) + 
  lims(x=c(0,0.75), y=c(0,0.75)) + 
  geom_hline(yintercept = 0, linetype="dashed", color="grey") +
  geom_vline(xintercept = 0, linetype="dashed", color="grey") +
  geom_point() + 
  geom_label_repel(max.overlaps = 100, seed = 1, ylim = c(0.1,2), force = 100, xlim = c(0,0.75), segment.linetype = "dashed") +
  theme_bw() + labs(x="ERV-W log2FC", y="Syncytin log2FC") +
  ggtitle("Correlation of log2FC in core enrichment genes (ERV-W GSEA)")
```

## Gene expression changes in relation to proximity to an ERV-W: Day 30 iNGNs

Does gene expression changes upon CRISPRa of ERV-W in neurons?

```{r}
path <- "./data/LTR17_distance_bin_day30/" # Output dir
dir.create(path)

protein_coding_genes <- fread("./data/gencode.v38.annotation_protein_coding_genes.bed", skip=1)
colnames(protein_coding_genes) <- c("chr", "start", "end", "gene", "dot", "strand")

# Set gene names as rownames
gene_res_list_df$Ca_LV4009_E01213_30dpi$gene <- rownames(gene_res_list_df$Ca_LV4009_E01213_30dpi)

# Extract all protein coding genes from the results
protein_coding_genes$log2FC <- gene_res_list_df$Ca_LV4009_E01213_30dpi[protein_coding_genes$gene, "log2FoldChange"]

# Make a bed to intersect to ERV-Ws
write.table(protein_coding_genes, "./data/LTR17_distance_bin_day30/gene_DEA_Ca_LV4009_E01213_30dpi_protein_coding.bed", sep="\t", quote = F, row.names = F, col.names = F)
```

## Intersection of protein coding gene promoters with ERV-Ws

1. Extract TSS of the genes
```{sh}
awk '{if($6=="+"){print $1,$2,$2+1,$4,$5,$6,$7}else{print $1,$3-1,$3,$4,$5,$6,$7}}' OFS="\t" ./data/LTR17_distance_bin_day30/gene_DEA_Ca_LV4009_E01213_30dpi_protein_coding.bed > ./data/LTR17_distance_bin_day30/gene_DEA_Ca_LV4009_E01213_30dpi_protein_coding_TSS.bed
```

2. Intersect with LTR17s (since this is the family we made the guides towards)

Within 1000 bps (`bedtools window -w 1000`), 5000 bps (`bedtools window -w 5000`), etc
```{sh}
bedtools window -a ./data/hg38_rmsk_TEtranscripts_LTR17.bed -b ./data/LTR17_distance_bin_day30/gene_DEA_Ca_LV4009_E01213_30dpi_protein_coding_TSS.bed -w 1000 -bed > ./data/LTR17_distance_bin_day30/gene_DEA_Ca_LV4009_E01213_30dpi_protein_coding_TSS_w1kbp_LTR17.bed

bedtools window -a ./data/hg38_rmsk_TEtranscripts_LTR17.bed -b ./data/LTR17_distance_bin_day30/gene_DEA_Ca_LV4009_E01213_30dpi_protein_coding_TSS.bed -w 5000 -bed > ./data/LTR17_distance_bin_day30/gene_DEA_Ca_LV4009_E01213_30dpi_protein_coding_TSS_w5kbp_LTR17.bed

bedtools window -a ./data/hg38_rmsk_TEtranscripts_LTR17.bed -b ./data/LTR17_distance_bin_day30/gene_DEA_Ca_LV4009_E01213_30dpi_protein_coding_TSS.bed -w 10000 -bed > ./data/LTR17_distance_bin_day30/gene_DEA_Ca_LV4009_E01213_30dpi_protein_coding_TSS_w10kbp_LTR17.bed

bedtools window -a ./data/hg38_rmsk_TEtranscripts_LTR17.bed -b ./data/LTR17_distance_bin_day30/gene_DEA_Ca_LV4009_E01213_30dpi_protein_coding_TSS.bed -w 15000 -bed > ./data/LTR17_distance_bin_day30/gene_DEA_Ca_LV4009_E01213_30dpi_protein_coding_TSS_w15kbp_LTR17.bed

bedtools window -a ./data/hg38_rmsk_TEtranscripts_LTR17.bed -b ./data/LTR17_distance_bin_day30/gene_DEA_Ca_LV4009_E01213_30dpi_protein_coding_TSS.bed -w 25000 -bed > ./data/LTR17_distance_bin_day30/gene_DEA_Ca_LV4009_E01213_30dpi_protein_coding_TSS_w25kbp_LTR17.bed

bedtools window -a ./data/hg38_rmsk_TEtranscripts_LTR17.bed -b ./data/LTR17_distance_bin_day30/gene_DEA_Ca_LV4009_E01213_30dpi_protein_coding_TSS.bed -w 50000 -bed > ./data/LTR17_distance_bin_day30/gene_DEA_Ca_LV4009_E01213_30dpi_protein_coding_TSS_w50kbp_LTR17.bed
```

### Visualize the results

Load the data
```{r}
files <- list.files(path)
files <- files[which(grepl("w1kbp_LTR17|w5kbp_LTR17|w10kbp_LTR17|w15kbp_LTR17|w25kbp_LTR17|w50kbp_LTR17", files))]
files

gene_DEA_30dpi_protein_coding_TSS_w1kbp <- fread(paste0(path, "/gene_DEA_Ca_LV4009_E01213_30dpi_protein_coding_TSS_w1kbp_LTR17.bed"))
gene_DEA_30dpi_protein_coding_TSS_w5kbp <- fread(paste0(path, "/gene_DEA_Ca_LV4009_E01213_30dpi_protein_coding_TSS_w5kbp_LTR17.bed"))
gene_DEA_30dpi_protein_coding_TSS_w10kbp <- fread(paste0(path, "/gene_DEA_Ca_LV4009_E01213_30dpi_protein_coding_TSS_w10kbp_LTR17.bed"))
gene_DEA_30dpi_protein_coding_TSS_w15kbp <- fread(paste0(path, "/gene_DEA_Ca_LV4009_E01213_30dpi_protein_coding_TSS_w15kbp_LTR17.bed"))
gene_DEA_30dpi_protein_coding_TSS_w25kbp <- fread(paste0(path, "/gene_DEA_Ca_LV4009_E01213_30dpi_protein_coding_TSS_w25kbp_LTR17.bed"))
gene_DEA_30dpi_protein_coding_TSS_w50kbp <- fread(paste0(path, "/gene_DEA_Ca_LV4009_E01213_30dpi_protein_coding_TSS_w50kbp_LTR17.bed"))

cols <- c("LTR_chr", "LTR_start", "LTR_end", "LTR_id", "LTR_dot", "LTR_strand", "LTR_info", "gene_chr", "gene_TSS_start", "gene_TSS_end", "gene", "gene_dot", "gene_strand", "gene_log2FC")

colnames(gene_DEA_30dpi_protein_coding_TSS_w1kbp) <- cols
colnames(gene_DEA_30dpi_protein_coding_TSS_w5kbp) <- cols
colnames(gene_DEA_30dpi_protein_coding_TSS_w10kbp) <- cols
colnames(gene_DEA_30dpi_protein_coding_TSS_w15kbp) <- cols
colnames(gene_DEA_30dpi_protein_coding_TSS_w25kbp) <- cols
colnames(gene_DEA_30dpi_protein_coding_TSS_w50kbp) <- cols
```


Remove duplicated entries. Namely, if a gene is in the 1kbp range, it will be in all the others too, so we want to keep it only in the closest window (1kbp). 
```{r}
gene_DEA_30dpi_protein_coding_TSS_w50kbp <- gene_DEA_30dpi_protein_coding_TSS_w50kbp[which(!gene_DEA_30dpi_protein_coding_TSS_w50kbp$gene %in% gene_DEA_30dpi_protein_coding_TSS_w25kbp$gene),]
gene_DEA_30dpi_protein_coding_TSS_w25kbp <- gene_DEA_30dpi_protein_coding_TSS_w25kbp[which(!gene_DEA_30dpi_protein_coding_TSS_w25kbp$gene %in% gene_DEA_30dpi_protein_coding_TSS_w15kbp$gene),]
gene_DEA_30dpi_protein_coding_TSS_w15kbp <- gene_DEA_30dpi_protein_coding_TSS_w15kbp[which(!gene_DEA_30dpi_protein_coding_TSS_w15kbp$gene %in% gene_DEA_30dpi_protein_coding_TSS_w10kbp$gene),]
gene_DEA_30dpi_protein_coding_TSS_w10kbp <- gene_DEA_30dpi_protein_coding_TSS_w10kbp[which(!gene_DEA_30dpi_protein_coding_TSS_w10kbp$gene %in% gene_DEA_30dpi_protein_coding_TSS_w5kbp$gene),]
gene_DEA_30dpi_protein_coding_TSS_w5kbp <- gene_DEA_30dpi_protein_coding_TSS_w5kbp[which(!gene_DEA_30dpi_protein_coding_TSS_w5kbp$gene %in% gene_DEA_30dpi_protein_coding_TSS_w1kbp$gene),]

gene_DEA_30dpi_protein_coding_TSS_w1kbp$distance <- "w1kbp"
gene_DEA_30dpi_protein_coding_TSS_w5kbp$distance <- "w5kbp"
gene_DEA_30dpi_protein_coding_TSS_w10kbp$distance <- "w10kbp"
gene_DEA_30dpi_protein_coding_TSS_w15kbp$distance <- "w15kbp"
gene_DEA_30dpi_protein_coding_TSS_w25kbp$distance <- "w25kbp"
gene_DEA_30dpi_protein_coding_TSS_w50kbp$distance <- "w50kbp"

gene_DEA_30dpi_protein_coding_TSS <- rbind(gene_DEA_30dpi_protein_coding_TSS_w1kbp,
                                           gene_DEA_30dpi_protein_coding_TSS_w5kbp,
                                           gene_DEA_30dpi_protein_coding_TSS_w10kbp,
                                           gene_DEA_30dpi_protein_coding_TSS_w15kbp,
                                           gene_DEA_30dpi_protein_coding_TSS_w25kbp,
                                           gene_DEA_30dpi_protein_coding_TSS_w50kbp)

gene_DEA_30dpi_protein_coding_TSS$distance <- factor(gene_DEA_30dpi_protein_coding_TSS$distance, levels = c("w1kbp", "w5kbp", "w10kbp", "w15kbp", "w25kbp", "w50kbp"))
```

### Now we plot! Day 30 iNGN gene effect in relation to ERV-W LTR proximity
```{r}
gene_DEA_30dpi_protein_coding_TSS %>% 
  ggplot(aes(x=distance, y=gene_log2FC)) + 
  geom_boxplot(outliers = F) +
  geom_jitter(height = 0, width = 0.2, size=0.5) + 
  theme_bw() + 
  labs(x="Distance to gene TSS", y="Gene log2FC") + ggtitle("ERV-W CRISPRa day 30 iNGNs", subtitle = "Protein coding only")
```

## Gene expression changes in relation to proximity to an ERV-W: hiPSC

Does gene expression changes upon CRISPRa of ERV-W in hiPSC?
```{r}
path <- "./data/LTR17_distance_bin_hiPSC/" # Output dir
dir.create(path)
# We load again because we used that dataframe to add the log2FC before. We will do the same again but with the hiPSC results object
protein_coding_genes <- fread("./data/gencode.v38.annotation_protein_coding_genes.bed", skip=1)
colnames(protein_coding_genes) <- c("chr", "start", "end", "gene", "dot", "strand")

gene_res_list_df$Ca_LV4009_E0010405061213_0dpi$gene <- rownames(gene_res_list_df$Ca_LV4009_E0010405061213_0dpi)

protein_coding_genes$log2FC <- gene_res_list_df$Ca_LV4009_E0010405061213_0dpi[protein_coding_genes$gene, "log2FoldChange"]

write.table(protein_coding_genes, "./data/LTR17_distance_bin_hiPSC/gene_DEA_Ca_LV4009_E0010405061213_0dpi_protein_coding.bed", sep="\t", quote = F, row.names = F, col.names = F)
```

## Intersection of protein coding gene promoters with ERV-Ws

1. Extract TSS of the genes
```{sh}
awk '{if($6=="+"){print $1,$2,$2+1,$4,$5,$6,$7}else{print $1,$3-1,$3,$4,$5,$6,$7}}' OFS="\t" ./data/LTR17_distance_bin_hiPSC/gene_DEA_Ca_LV4009_E0010405061213_0dpi_protein_coding.bed > ./data/LTR17_distance_bin_hiPSC/gene_DEA_Ca_LV4009_E0010405061213_0dpi_protein_coding_TSS.bed
```

2. Intersect with LTR17s (since this is the family we made the guides towards)

Within 1000 bps (`bedtools window -w 1000`), 5000 bps (`bedtools window -w 5000`), etc
```{sh}
bedtools window -a ./data/hg38_rmsk_TEtranscripts_LTR17.bed -b ./data/LTR17_distance_bin_hiPSC/gene_DEA_Ca_LV4009_E0010405061213_0dpi_protein_coding_TSS.bed -w 1000 -bed > ./data/LTR17_distance_bin_hiPSC/gene_DEA_Ca_LV4009_E0010405061213_0dpi_protein_coding_TSS_w1kbp_LTR17.bed

bedtools window -a ./data/hg38_rmsk_TEtranscripts_LTR17.bed -b ./data/LTR17_distance_bin_hiPSC/gene_DEA_Ca_LV4009_E0010405061213_0dpi_protein_coding_TSS.bed -w 5000 -bed > ./data/LTR17_distance_bin_hiPSC/gene_DEA_Ca_LV4009_E0010405061213_0dpi_protein_coding_TSS_w5kbp_LTR17.bed

bedtools window -a ./data/hg38_rmsk_TEtranscripts_LTR17.bed -b ./data/LTR17_distance_bin_hiPSC/gene_DEA_Ca_LV4009_E0010405061213_0dpi_protein_coding_TSS.bed -w 10000 -bed > ./data/LTR17_distance_bin_hiPSC/gene_DEA_Ca_LV4009_E0010405061213_0dpi_protein_coding_TSS_w10kbp_LTR17.bed

bedtools window -a ./data/hg38_rmsk_TEtranscripts_LTR17.bed -b ./data/LTR17_distance_bin_hiPSC/gene_DEA_Ca_LV4009_E0010405061213_0dpi_protein_coding_TSS.bed -w 15000 -bed > ./data/LTR17_distance_bin_hiPSC/gene_DEA_Ca_LV4009_E0010405061213_0dpi_protein_coding_TSS_w15kbp_LTR17.bed

bedtools window -a ./data/hg38_rmsk_TEtranscripts_LTR17.bed -b ./data/LTR17_distance_bin_hiPSC/gene_DEA_Ca_LV4009_E0010405061213_0dpi_protein_coding_TSS.bed -w 25000 -bed > ./data/LTR17_distance_bin_hiPSC/gene_DEA_Ca_LV4009_E0010405061213_0dpi_protein_coding_TSS_w25kbp_LTR17.bed

bedtools window -a ./data/hg38_rmsk_TEtranscripts_LTR17.bed -b ./data/LTR17_distance_bin_hiPSC/gene_DEA_Ca_LV4009_E0010405061213_0dpi_protein_coding_TSS.bed -w 50000 -bed > ./data/LTR17_distance_bin_hiPSC/gene_DEA_Ca_LV4009_E0010405061213_0dpi_protein_coding_TSS_w50kbp_LTR17.bed
```

### Visualize the results

Load the data
```{r}
files <- list.files(path)
files <- files[which(grepl("w1kbp_LTR17|w5kbp_LTR17|w10kbp_LTR17|w15kbp_LTR17|w25kbp_LTR17|w50kbp_LTR17", files))]

gene_DEA_hipsc_protein_coding_TSS_w1kbp <- fread(paste0(path, "/gene_DEA_Ca_LV4009_E0010405061213_0dpi_protein_coding_TSS_w1kbp_LTR17.bed"))
gene_DEA_hipsc_protein_coding_TSS_w5kbp <- fread(paste0(path, "/gene_DEA_Ca_LV4009_E0010405061213_0dpi_protein_coding_TSS_w5kbp_LTR17.bed"))
gene_DEA_hipsc_protein_coding_TSS_w10kbp <- fread(paste0(path, "/gene_DEA_Ca_LV4009_E0010405061213_0dpi_protein_coding_TSS_w10kbp_LTR17.bed"))
gene_DEA_hipsc_protein_coding_TSS_w15kbp <- fread(paste0(path, "/gene_DEA_Ca_LV4009_E0010405061213_0dpi_protein_coding_TSS_w15kbp_LTR17.bed"))
gene_DEA_hipsc_protein_coding_TSS_w25kbp <- fread(paste0(path, "/gene_DEA_Ca_LV4009_E0010405061213_0dpi_protein_coding_TSS_w25kbp_LTR17.bed"))
gene_DEA_hipsc_protein_coding_TSS_w50kbp <- fread(paste0(path, "/gene_DEA_Ca_LV4009_E0010405061213_0dpi_protein_coding_TSS_w50kbp_LTR17.bed"))

cols <- c("LTR_chr", "LTR_start", "LTR_end", "LTR_id", "LTR_dot", "LTR_strand", "LTR_info", "gene_chr", "gene_TSS_start", "gene_TSS_end", "gene", "gene_dot", "gene_strand", "gene_log2FC")

colnames(gene_DEA_hipsc_protein_coding_TSS_w1kbp) <- cols
colnames(gene_DEA_hipsc_protein_coding_TSS_w5kbp) <- cols
colnames(gene_DEA_hipsc_protein_coding_TSS_w10kbp) <- cols
colnames(gene_DEA_hipsc_protein_coding_TSS_w15kbp) <- cols
colnames(gene_DEA_hipsc_protein_coding_TSS_w25kbp) <- cols
colnames(gene_DEA_hipsc_protein_coding_TSS_w50kbp) <- cols
```


Remove duplicated entries. Namely, if a gene is in the 1kbp range, it will be in all the others too, so we want to keep it only in the closest window (1kbp). 
```{r}
gene_DEA_hipsc_protein_coding_TSS_w50kbp <- gene_DEA_hipsc_protein_coding_TSS_w50kbp[which(!gene_DEA_hipsc_protein_coding_TSS_w50kbp$gene %in% gene_DEA_hipsc_protein_coding_TSS_w25kbp$gene),]
gene_DEA_hipsc_protein_coding_TSS_w25kbp <- gene_DEA_hipsc_protein_coding_TSS_w25kbp[which(!gene_DEA_hipsc_protein_coding_TSS_w25kbp$gene %in% gene_DEA_hipsc_protein_coding_TSS_w15kbp$gene),]
gene_DEA_hipsc_protein_coding_TSS_w15kbp <- gene_DEA_hipsc_protein_coding_TSS_w15kbp[which(!gene_DEA_hipsc_protein_coding_TSS_w15kbp$gene %in% gene_DEA_hipsc_protein_coding_TSS_w10kbp$gene),]
gene_DEA_hipsc_protein_coding_TSS_w10kbp <- gene_DEA_hipsc_protein_coding_TSS_w10kbp[which(!gene_DEA_hipsc_protein_coding_TSS_w10kbp$gene %in% gene_DEA_hipsc_protein_coding_TSS_w5kbp$gene),]
gene_DEA_hipsc_protein_coding_TSS_w5kbp <- gene_DEA_hipsc_protein_coding_TSS_w5kbp[which(!gene_DEA_hipsc_protein_coding_TSS_w5kbp$gene %in% gene_DEA_hipsc_protein_coding_TSS_w1kbp$gene),]

gene_DEA_hipsc_protein_coding_TSS_w1kbp$distance <- "w1kbp"
gene_DEA_hipsc_protein_coding_TSS_w5kbp$distance <- "w5kbp"
gene_DEA_hipsc_protein_coding_TSS_w10kbp$distance <- "w10kbp"
gene_DEA_hipsc_protein_coding_TSS_w15kbp$distance <- "w15kbp"
gene_DEA_hipsc_protein_coding_TSS_w25kbp$distance <- "w25kbp"
gene_DEA_hipsc_protein_coding_TSS_w50kbp$distance <- "w50kbp"

gene_DEA_hipsc_protein_coding_TSS <- rbind(gene_DEA_hipsc_protein_coding_TSS_w1kbp,
                                           gene_DEA_hipsc_protein_coding_TSS_w5kbp,
                                           gene_DEA_hipsc_protein_coding_TSS_w10kbp,
                                           gene_DEA_hipsc_protein_coding_TSS_w15kbp,
                                           gene_DEA_hipsc_protein_coding_TSS_w25kbp,
                                           gene_DEA_hipsc_protein_coding_TSS_w50kbp)

gene_DEA_hipsc_protein_coding_TSS$distance <- factor(gene_DEA_hipsc_protein_coding_TSS$distance, levels = c("w1kbp", "w5kbp", "w10kbp", "w15kbp", "w25kbp", "w50kbp"))
```

### Now we plot! hiPSC gene effect in relation to ERV-W LTR proximity
```{r}
gene_DEA_hipsc_protein_coding_TSS %>% 
  ggplot(aes(x=distance, y=gene_log2FC)) + 
  geom_boxplot(outliers = F) +
  geom_jitter(height = 0, width = 0.2, size=0.5) + 
  theme_bw() + 
  labs(x="Distance to gene TSS", y="Gene log2FC") + ggtitle("ERV-W CRISPRa hiPSC")

gene_DEA_hipsc_protein_coding_TSS %>% 
  filter(gene_DEA_hipsc_protein_coding_TSS$gene_log2FC > 2)
```
