---
title: "Single nuclei RNAseq results from day 32 iNGN sample"
output:
  html_document:
    df_print: paged
---

This is a check up of the cell type composition of our iNGN cultures at day 32.

### Load libraries and data
```{r}
library(Seurat)
library(ggplot2)
library(tidyverse)
library(data.table)

sample.data <- Read10X(data.dir = "~/inbox/MN11_HERVW/MN11_iNGN_Rep1_10x_32dpi_E015/outs/filtered_feature_bc_matrix/")
sample <- CreateSeuratObject(counts = sample.data, project = "iNGN", min.cells = 3, min.features = 500)
sample[["percent.mt"]] <- PercentageFeatureSet(sample, pattern = "^MT-")
```

### Quality control metrics and filtering
```{r}
# Before
VlnPlot(sample, c("nFeature_RNA", "percent.mt", "nCount_RNA"))

max_genes <- mean(sample$nFeature_RNA) + 2*(sd(sample$nFeature_RNA)) 
min_genes <- mean(sample$nFeature_RNA) - (sd(sample$nFeature_RNA))
min_genes <- ifelse(min_genes < 0, 0, min_genes)

sample_postqc <- subset(sample, subset = nFeature_RNA > min_genes & nFeature_RNA < max_genes)
# After filtering
VlnPlot(sample_postqc, c("nFeature_RNA", "percent.mt", "nCount_RNA"))
```

### Normalization, scaling, PCA and UMAP
```{r}
DefaultAssay(sample_postqc) <- "RNA"  
sample_postqc <- NormalizeData(sample_postqc, assay = "RNA", margin = 2) 
sample_postqc <- FindVariableFeatures(sample_postqc, selection.method = "vst", nfeatures = 2000)
all.genes <- rownames(sample_postqc)
sample_postqc <- RenameCells(sample_postqc, sample_postqc$orig.ident)

sample_postqc <- ScaleData(sample_postqc, features = all.genes)
sample_postqc <- RunPCA(sample_postqc, features = VariableFeatures(object = sample_postqc))
sample_postqc <- FindNeighbors(sample_postqc, dims = 1:10)
sample_postqc <- FindClusters(sample_postqc, resolution = 0.1)
sample_postqc <- RunUMAP(sample_postqc, dims = 1:10)

DimPlot(sample_postqc) & theme_bw()
```

### How many cells are NeuN+?
```{r}
FetchData(sample_postqc, c("RBFOX3", "umap_1", "umap_2")) %>% 
  arrange(RBFOX3) %>% 
  ggplot(aes(x=umap_1, y=umap_2, color=RBFOX3)) + geom_point(size=0.3) + theme_bw() +
  scale_color_gradient(low="lightgrey", high="red")
```

Interesting, does this relate to the number of genes detected? (did we set the original threshold too low?)
```{r}
FeaturePlot(sample_postqc, "nFeature_RNA") & theme_bw()
```

Seems like it, but let's see if we can find any genes marking this cells
```{r}
gene_dea <- FindAllMarkers(sample_postqc)
# Protein coding
protein_coding_genes <- fread("/Volumes/MyPassport/annotations/human/gencode/v38/gencode.v38.annotation_protein_coding_genes.bed", skip=1)

gene_dea_top10 <- gene_dea %>% 
  filter(avg_log2FC > 1 & pct.1 > 0.2 & gene %in% protein_coding_genes$V4) %>% 
  group_by(cluster) %>% 
  arrange(avg_log2FC, -p_val_adj) %>% 
  top_n(10, avg_log2FC) %>% 
  arrange(cluster)
  
DotPlot(sample_postqc, unique(gene_dea_top10$gene)) + coord_flip()
```

Ok, seems like all markers of this cluster are also expressed in the rest of the clusters (to a lower intensity), and are expressed in a lower fraction of the cells (as a percentage of the cluster). *This looks like fragmented cells, I'll exclude them.*

## New QC - minimum of 1000 genes detected

```{r}
sample <- CreateSeuratObject(counts = sample.data, project = "iNGN", min.cells = 3, min.features = 1000)
sample[["percent.mt"]] <- PercentageFeatureSet(sample, pattern = "^MT-")
# Before the std dev filtering
VlnPlot(sample, c("nFeature_RNA", "percent.mt", "nCount_RNA"))

max_genes <- mean(sample$nFeature_RNA) + 2*(sd(sample$nFeature_RNA)) 
min_genes <- mean(sample$nFeature_RNA) - (sd(sample$nFeature_RNA))
min_genes <- ifelse(min_genes < 0, 0, min_genes)

sample_postqc <- subset(sample, subset = nFeature_RNA > min_genes & nFeature_RNA < max_genes)
# After filtering
VlnPlot(sample_postqc, c("nFeature_RNA", "percent.mt", "nCount_RNA"))
sample_postqc
```

### Normalize, scale, PCA and UMAP

```{r}
DefaultAssay(sample_postqc) <- "RNA"  
sample_postqc <- NormalizeData(sample_postqc, assay = "RNA", margin = 2) 
sample_postqc <- FindVariableFeatures(sample_postqc, selection.method = "vst", nfeatures = 2000)
all.genes <- rownames(sample_postqc)
sample_postqc <- RenameCells(sample_postqc, sample_postqc$orig.ident)

sample_postqc <- ScaleData(sample_postqc, features = all.genes)
sample_postqc <- RunPCA(sample_postqc, features = VariableFeatures(object = sample_postqc))
sample_postqc <- FindNeighbors(sample_postqc, dims = 1:10)
sample_postqc <- FindClusters(sample_postqc, resolution = 0.1)
sample_postqc <- RunUMAP(sample_postqc, dims = 1:10)
DimPlot(sample_postqc) & theme_bw()
```

### How many cells are NeuN+?
```{r}
FetchData(sample_postqc, c("RBFOX3", "umap_1", "umap_2")) %>% 
  arrange(RBFOX3) %>% 
  ggplot(aes(x=umap_1, y=umap_2, color=RBFOX3)) + geom_point(size=0.3) + theme_bw() +
  scale_color_gradient(low="lightgrey", high="red")
```

### What type of neurons?
```{r}
gene_dea <- FindAllMarkers(sample_postqc)
gene_dea_top10 <- gene_dea %>% 
  filter(avg_log2FC > 1 & pct.1 > 0.2 & gene %in% protein_coding_genes$V4) %>% 
  group_by(cluster) %>% 
  arrange(avg_log2FC, -p_val_adj) %>% 
  top_n(10, avg_log2FC) %>% 
  arrange(cluster)

DotPlot(sample_postqc, unique(gene_dea_top10$gene)) + coord_flip()
```

